Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-12 ms.tex
editing

Radiative cooling, if important, qualitative alters key features of the accretion flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm cool} = n^{2}\Lambda \propto T^{-2.7}$ (see below) is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \in [-0.7,0.5]$.  If manifest, thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Wang+14}).

@@

@@

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{\rm w,0} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{\rm w,0}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon = 1$.  On the other hand, the
binaries giving rise to MSPs may be disocciated by stellar
interactions in the dense nuclear cluster, reducing their numbers as
compared to the field population estimate above.  We conclude that
MSPs are unlikely to be an important source of heating, probably
insufficient to stabilize the steady flow against cooling instability
(e.g. $v_{\rm w,0}^{\rm MSP} < v_{\rm critt}$;
eq.~[\ref{eq:cooling2}])

Type Ia supernovae (SNe Ia) represent a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  If each Ia supernova injects thermal energy $E_{\rm Ia}$ into the interstellar medium, and the SNe rate (per stellar mass) is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm  Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}]) \begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.

The high value of $v_{w,0}^{\rm Ia}$ implies that Ia SNe represent a potentially dominant source of non-AGN heating in the CNM.  However, SNe can only be approximated as being a constant heating source throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship
(equation~[\ref{eq:Msigma}]).

What is the effect of the Ia radius?  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \approx M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{\rm w,0}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence Ia SN can in principle periodically clear out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.

>>>>

Radiative cooling, if important, qualitative alters key features of the accretion flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm cool} = n^{2}\Lambda \propto T^{\beta-2}$ (see below) is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \in [-0.7,0.5]$.  If manifest, thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Wang+14}).

@@

Type Ia supernovae (SNe Ia) represent a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  If each Ia supernova injects thermal energy $E_{\rm Ia}$ into the interstellar medium, and the SNe rate (per stellar mass) is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm  Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}]) \begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.  We hereafter adopt $\epsilon_{\rm Ia} = 0.4$.

The high value of $v_{w,0}^{\rm Ia}$ implies that Ia SNe represent an important source of CNM heating.  However, SNe can only be approximated as supplying constant heating throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship
(equation~[\ref{eq:Msigma}]).

What is the effect of the Ia radius?  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \approx M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{\rm w,0}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence Ia SN can in principle dynamically clear out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Hence, even if $r_{\rm s} > r_{\rm Ia}$, the accretion rate will be limited to a value set by substituting $r_{\rm Ia}$ for $r_{\rm s}$ in derivigin equation $\ref{eq:eddr_analytic}$, i.e.  {\bf calculate:}
\begin{eqnarray}
\left.\frac{\dot{M}}{\dot{M}_{\rm edd}}\right|_{\rm max} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}

@@

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{\rm w,0} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{\rm w,0}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon = 1$.  On the other hand, the
binaries giving rise to MSPs may be disocciated by stellar
interactions in the dense nuclear cluster, reducing their numbers as
compared to the field population estimate above.  MSPs are thus unlikely to be an important source of heating in comparison to stellar heating.
